---
title: "Getting Started with phyloPal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phyloPal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(bitmapType = "cairo")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 96,
  out.width = "100%",
  dev = "CairoPNG"
)
```

```{r setup, message = FALSE}
library(phyloPal)
library(dplyr)
library(ggplot2) 
library(cowplot) 
library(ggplotify)
```

## Overview

phyloPal is designed around a common bottleneck in microbiome data visualization: 
the gap between having processed abundance data and producing figures that are both 
publication-ready and honest about  compositional structure. 
The package provides a connected workflow — taxonomy cleaning, data aggregation, 
color palette generation, and plotting — where each step feeds naturally into the next. 
Palettes are aware of taxonomic hierarchy, alluvial plots are aware of which taxa are shared 
versus unique across groups, and combined dendrogram-alluvial figures keep beta diversity and 
composition aligned in the same panel. 

The examples in this vignette use a subset of the `GlobalPatterns` dataset 
from the `phyloseq` package, covering five habitat types: 
Terrestrial, Oceanic, Freshwater, Brackish, and Freshwater creek.

## Installation

phyloPal is available from GitHub. The `devtools` package is required for installation:

```{R installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mwslawinska/phyloPal")
```

## Example data

phyloPal ships with a subset of the `GlobalPatterns` dataset from the
`phyloseq` package, filtered to five habitat types.

```{r data}
data(example_microbiome)
data(em_metadata)
data(em_otu)

# What does it look like?
glimpse(example_microbiome) # long-format ASV table with RA column
glimpse(em_metadata) # sample metadata
glimpse(em_otu)
```

## Workflows

### 1. Data preparation and taxonomy cleaning

phyloPal works with long-format ASV/OTU tables where RA is already calculated per ASV/OTU.

A built-in taxonomy cleaner `replace_incertae_sedis_NAs()` standardizes hierarchical taxonomy columns by 
normalizing common "Incertae sedis" variants (e.g. "Incertae_Sedis", "incertae sedis"), 
filling missing child ranks from stable parent ranks (e.g. a missing Family becomes "Rhizobiales, unclassified" if Order is known), 
and replacing empty or uninformative entries with `"unknown"`.

All other phyloPal functions call this cleaner internally by default 
(`clean_taxonomy = TRUE`), so explicit pre-cleaning is optional but 
recommended when you want full control over the taxonomy before 
aggregation.
```{R clean}
em_cleaned <- example_microbiome %>%
replace_incertae_sedis_NAs()

glimpse(em_cleaned)

```

### 2. Data aggregation and color palettes
Before plotting barplots, raw ASV-level data must be aggregated to the desired taxonomic level. 
`process_barplot_data()` handles aggregation, normalization, and low-abundance grouping in one step.

#### Low-abundance handling
A key decision is how to handle low-abundance taxa — those below `low_abundance_threshold`. 
The `keep_ratype` argument controls this:
- **`"collapse"`** (simpler): all taxa below the threshold are relabelled as "low abundant" and merged into a single bin. 
This keeps the plot clean and is the right choice when you only care about the dominant taxa.
- **`"separate"`** (more flexible): low-abundance taxa are flagged but 
their original identity is preserved in a `<tax_level>_original` column. 
The plot-level label becomes `"low abundant"`, but the true taxon name is retained for downstream use.

The `low_abundance_basis` argument controls **when** the threshold is 
applied:
- **`"per_sample"`**: taxa are flagged as low abundant within each individual sample before aggregation. 
- **`"post_aggregation"`**: the threshold is applied after averaging across samples or groups. 

#### Aggregation function
The `agg_fun` argument controls how relative abundances are combined when multiple ASVs map to the same taxon within a sample: 
- **`"sum"`** adds them together, giving the total relative abundance of that taxon in the sample. 
- **`"mean"`** instead averages across ASVs, which is rarely wanted at the within-sample level 
but can make sense in specific workflows.
In most cases, use `agg_fun = "sum"`.

### Sample-level aggregation

```{R barplot-data}
em_barplot_processed <- process_barplot_data(
  em_cleaned,
  tax_level = "Class",
  group_vars = c("SampleType", "SampleID", "Habitat"),
  low_abundance_basis = "per_sample",
  low_abundance_threshold = 0.01,
  agg_fun = "sum",
  keep_ratype = "separate",
  clean_taxonomy = FALSE
)


em_barplot_processed2 <- process_barplot_data(
  example_microbiome,
tax_level = "Class",
group_vars = c("SampleType", "SampleID", "Habitat"),
low_abundance_threshold = 0.01,
keep_ratype = "collapse",
clean_taxonomy = TRUE,
preserve_higher_taxonomy = T
)

glimpse(em_barplot_processed)
glimpse(em_barplot_processed2)
```

#### Group-level aggregation and barplot

Rather than keeping individual samples, data can be aggregated to the 
group level — for example, averaging relative abundances across all 
samples of the same `SampleType`. This is useful when you have many 
samples per group and want a single representative bar per group, or 
when you want to compare broad habitat-level patterns rather than 
sample-level variation. To do this, pass the grouping variables to 
`group_vars` and set `normalize_by` to the same variables — this tells 
`process_barplot_data()` to normalize within groups rather than within 
individual samples. The resulting data frame has one row per taxon per 
group, ready for plotting with `plot_taxonomic_barplot()`.

```{R process-barplot-grouped}
em_processed_grouped2 <- process_barplot_data(
  example_microbiome,
  tax_level = "Class",
  group_vars = c("SampleType", "Habitat"),
  normalize_by = c("SampleType", "Habitat"),
  low_abundance_threshold = 0.01,
  preserve_higher_taxonomy = TRUE,
  low_abundance_basis = "per_sample",
  agg_fun = "sum",
  keep_ratype = "separate"
)
```

#### Color palettes

`generate_palette_hcl()` generates perceptually uniform HCL palettes for any number of taxa. 
HCL (Hue-Chroma-Luminance) colors are preferred for scientific visualization 
because equal steps in HCL space correspond to equal perceived differences in color — 
unlike RGB-based palettes where some colors appear much brighter or more saturated than others.


```{R barplot-palette-simple}
barplot_pal <- generate_palette_hcl(
  data = em_barplot_processed,
  tax_level = "Class",
  fixed_colors_enabled = TRUE,
  fixed_colors_position = "end",
  palette_list = c("Reds", "Purples", "BrwnYl", "Blues", "TealGrn"),
  cmax = 65,
  luminance = c(20,90),
    power = 1.2,
    shuffle = FALSE)
```

#### Grouped palette for sample metadata

`generate_grouped_palette()` assigns colors from the same color family 
to items sharing a higher-level group — for example, all freshwater 
sample types in green tones, all oceanic sample types in blue tones. 
Beyond facet strip coloring, this palette is useful whenever 
consistent color coding across multiple figure types is needed: 
if barplot facet strips, dendrogram labels, and beta-diversity ordination 
plots are colored by the same habitat palette, all figures in a panel share a common 
visual language and the reader only needs to learn the color scheme once. 
This is particularly valuable when plotting multiple samples per group — 
for example, PCoA or NMDS plots where individual samples are colored by 
their higher-level group membership.

```{R grouped-palette}

habitat_palette <- generate_grouped_palette(
  data = em_cleaned,
  group_col = "Habitat",
  item_col = "SampleType",
  palette_map = list(
    "Terrestrial" = "BrwnYl",
    "Oceanic" = "Blues",
    "Freshwater" = "Greens",
    "Brackish" = "PuRd"
  ),
  luminance = 65,
  power = 1.2
)
```

#### Grouped palette for taxa
The principle used in `generate_grouped_palette()` can be applied to taxonomic palettes, too.
When datasets contain many taxa, a flat palette where colors are 
assigned arbitrarily can make it hard to orient visually. 
`generate_palette_hcl()` supports optional hierarchical grouping via 
`group_by_higher_tax` and `group_palette_map` — all families belonging 
to Proteobacteria get blue tones, all Actinobacteria get red tones, and 
so on. This makes the biological structure of the community visible at 
a glance rather than requiring careful legend inspection. 
However, hierarchical grouping is recommended only for smaller datasets with 
fewer than ~10 higher-level taxa — synthetic communities are a typical 
use case. For complex natural communities with many phyla or classes, 
the ungrouped palette is typically more interpretable, as too many 
color families become difficult to distinguish.

```{R barplot-palette-grouped}
em_processed_grouped <- process_barplot_data(
  example_microbiome,
  tax_level = "Class",
  group_vars = c("SampleType", "Habitat"),
  normalize_by = c("SampleType", "Habitat"),
  low_abundance_threshold = 0.1,
  preserve_higher_taxonomy = TRUE,
  low_abundance_basis = "per_sample",
  agg_fun = "sum",
  keep_ratype = "separate"
)

barplot_pal_grouped <- generate_palette_hcl(
  data = em_processed_grouped,
  tax_level = "Class",
   group_by_higher_tax = "Phylum",
   order_by_higher_tax = TRUE,
   group_palette_map = list(
     "Actinobacteria" = "Blues",
     "Proteobacteria" = "Greens",
     "Cyanobacteria" = list(palette = "Purple-Orange", side = "right"),
     "Acidobacteria" = list(palette = "Purple-Orange", side = "right"),
     "Bacteroidetes" = "Burg",
     "Verrucomicrobia" = "BrwnYl"),
  fixed_colors_enabled = TRUE,
  fixed_colors_position = "end",
  order_groups = "alphabetical",
  order_within_groups = "alphabetical",
  cmax = 65,
  luminance = c(20,90),
    power = 1.2,
    shuffle = FALSE)
```

### 3. Taxonomic barplots

#### Colored facet strips

Facet strips — the label bars above or beside each panel in a faceted 
plot — are a missed opportunity in most microbiome figures. By default 
they carry only a text label, but coloring them to match a higher-level 
grouping variable adds a second layer of information without cluttering 
the plot. For example, if samples are faceted by `SampleType` but the `Habitat` 
each sample type belongs to should also be visible, coloring the strips 
by habitat lets the reader group panels visually — all freshwater panels 
share one color, all oceanic panels another — without adding an extra 
legend or annotation layer.

In base `ggplot2` and even with `ggh4x`, achieving this requires verbose 
boilerplate code that is easy to get wrong and hard to keep consistent 
across figures. In phyloPal, you pass a named color vector directly to 
`facet_strip_colors` in `plot_taxonomic_barplot()` — the same vector 
produced by `generate_grouped_palette()` — and the coloring is handled 
automatically. This makes it straightforward to keep strip colors 
consistent with other plot elements such as dendrogram labels, 
ordination point colors, or sample metadata annotations across all 
figures in a panel.

```{R barplot-separate}
p_barplot <- plot_taxonomic_barplot(
  data = em_barplot_processed,
  tax_level = "Class",
  palette = barplot_pal,
  x_axis_var = "SampleID",
  facet_by = "SampleType",
  facet_strip_colors = habitat_palette,
  theme_obj = theme_phylopal()
) + 
  ggplot2::guides(
    fill = guide_legend(
      ncol = 1
    )
  ) 

p_barplot
```

The two data processing approaches produce visually different results. 
With `keep_ratype = "separate"`, each low-abundance taxon retains its 
own identity in the `<tax_level>_original` column and is shown 
individually in the plot. With `keep_ratype = "collapse"`, all 
low-abundance taxa are merged into a single `"low abundant"` bin, 
producing a cleaner barplot. The choice depends on 
whether the identity of rare taxa matters for interpretation.

```{R barplot-collapse}
p_barplot2 <- plot_taxonomic_barplot(
  data = em_barplot_processed2,
  tax_level = "Class",
  palette = barplot_pal,
  x_axis_var = "SampleID",
  facet_by = "SampleType",
  facet_strip_colors = habitat_palette,
  theme_obj = theme_phylopal()
) + 
  ggplot2::guides(
    fill = guide_legend(
      ncol = 1
    )
  ) 

p_barplot2
```

#### Group-level barplot

Using the group-level aggregated data prepared in the previous section, 
`plot_taxonomic_barplot()` produces one bar per group rather than one 
bar per sample — useful for comparing broad habitat-level patterns 
across conditions.

```{R barplot-grouped-normal-palette}
p_barplot_grouped2 <- plot_taxonomic_barplot(
  data = em_processed_grouped2,
  tax_level = "Class",
  palette = barplot_pal,
  x_axis_var = "SampleType",
  facet_by = "Habitat",
  theme_obj = theme_phylopal() 
) + 
  ggplot2::guides(
    fill = guide_legend(
      ncol = 1
    )
  )  + theme(axis.text.x = element_text(size =11, angle = 45, hjust = 1, vjust = 1),
    axis.ticks.x = ggplot2::element_line(color = "black", linewidth = 0.4))

p_barplot_grouped2
```

#### Barplot with hierarchical taxonomic palette

The same group-level data can be plotted with a hierarchically grouped 
palette, where taxa belonging to the same phylum share a color family. 
This makes it easier to identify the dominant phylum in each habitat 
at a glance, without tracing every taxon back to the legend.

```{R barplot-grouped-palette}
p_barplot_grouped <- plot_taxonomic_barplot(
  data = em_processed_grouped,
  tax_level = "Class",
  palette = barplot_pal_grouped,
  x_axis_var = "SampleType",
  facet_by = "Habitat",
  theme_obj = theme_phylopal() 
) + 
  ggplot2::guides(
    fill = guide_legend(
      ncol = 1
    )
  )  + theme(axis.text.x = element_text(size =11, angle = 45, hjust = 1, vjust = 1),
    axis.ticks.x = ggplot2::element_line(color = "black", linewidth = 0.4))

p_barplot_grouped
```

### 4. Alluvial plots

Alluvial plots (also called Sankey diagrams) show how compositional structure changes across groups — which taxa are present in all groups, which are unique to one, and which shift in abundance between conditions. 

#### Taxa classification
Before plotting, taxa must be classified by their abundance pattern 
across groups using `classify_taxa_patterns()`, which assigns each 
taxon to one of four categories:
- **shared abundant**: present and abundant in all groups
- **shared low abundant**: present in all groups but always below the 
  threshold
- **unique abundant**: abundant in some groups but absent in others
- **unique low abundant**: present in only some groups and always rare

Taxa that are abundant in some groups but low in others are optionally 
detected as **shared mixed abundance** (enabled by default). These 
categories determine both the palette key assigned to each taxon and 
its stacking position in the plot — shared taxa appear at the bottom, 
unique taxa toward the top, and fixed categories like `"unknown"` and 
`"low abundant"` always occupy consistent positions.

#### Step-by-step workflow
The full workflow requires four steps:
`prepare_alluvial_data()` → `classify_taxa_patterns()` → 
`generate_alluvial_palette()` → `plot_alluvial()`.
Each step can be customised independently — for example, using a 
hierarchical grouped palette, passing special taxa that should never 
be collapsed into the low-abundance bin, or adjusting classification 
thresholds independently of the palette.

```{R alluvial-steps}
# arrange the SampleType like you want
example_microbiome$SampleType <- factor(example_microbiome$SampleType, 
levels = unique(example_microbiome$SampleType))

# prepare alluvial data
em_allu <- prepare_alluvial_data(example_microbiome,
tax_level = "Class",
group_col = c("SampleType"),
clean_taxonomy = TRUE
)

# classify taxa patterns according to their abundance
em_allu_classified <- classify_taxa_patterns(
  data = em_allu,
  tax_level = "Class",
  group_col = c("SampleType")
)

glimpse(em_allu_classified)

# generate palette for the alluvial plot
allu_pal <- generate_alluvial_palette(
    data = em_allu_classified,
  palette_list = c("Reds", "Purples", "BrwnYl", "Blues", "TealGrn"),
  cmax = 65,
  luminance = c(20,90),
    power = 1.2,
    )

# plot the alluvial plot
p_allu <- plot_alluvial(em_allu_classified, 
custom_palette = allu_pal,
tax_level = "Class", 
group_col = "SampleType",
theme_obj = theme_phylopal(),
line_width = 0.2,
x_axis_label = "Sample Type"
) +
ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggplot2::guides(
    fill = guide_legend(
      ncol = 1
    )
  )

p_allu
```

#### Convenience wrapper

For standard use cases, `create_alluvial_plot()` wraps the entire 
workflow into a single call, accepting nested argument lists 
(`prepare_args`, `classify_args`, `palette_args`, `plot_args`) passed 
through to each step. This makes it straightforward to go from raw 
data to a finished plot in a few lines, while retaining the option to 
drop into the step-by-step workflow whenever more control is needed.

```{R alluvial-wrapper}
em_allu_wrapper <- create_alluvial_plot(
  data = example_microbiome,
  tax_level = "Class",
  group_col = "SampleType",
  prepare_args = list(clean_taxonomy = TRUE),
  palette_list = c("Reds", "Purples", "BrwnYl", "Blues", "TealGrn"),
  palette_args = list(
    cmax = 65,
    luminance = c(20, 90),
    power = 1.2
  ),
  plot_args = list(
    theme_obj = theme_phylopal(),
    line_width = 0.2,
    x_axis_label = "Sample Type"
  )
) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggplot2::guides(fill = ggplot2::guide_legend(ncol = 1))

em_allu_wrapper
```
### 5. Alluvial plot combined with dendrogram

While an alluvial plot shows taxonomic composition across groups, it 
carries no information about how similar those groups are to each other 
overall. A sample dendrogram addresses this — it clusters groups by 
beta diversity (here Bray-Curtis dissimilarity) and reveals which 
communities are most similar in overall structure. Combining both plots 
in a single figure lets the reader interpret compositional patterns in 
the context of community-level relationships: groups that cluster 
closely in the dendrogram are expected to share more taxa in the 
alluvial plot, and deviations from this expectation become immediately 
visible.

#### Building the dendrogram

To create a dendrogram from grouped data (one column per `SampleType` 
rather than per sample), first average the ASV/OTU matrix within groups 
using `create_grouped_matrix()`, then build the dendrogram with 
`build_dendrogram()` and plot it with `plot_dendrogram()`. The 
`color_by` and `shape_by` arguments in `plot_dendrogram()` allow 
metadata variables to be encoded directly on the dendrogram labels, 
keeping it visually consistent with the alluvial plot and other 
figures in the panel.

Beta-diversity distances are computed using the `vegan` package 
(Oksanen et al., 2022).

```{r dendrogram, fig.height = 4}
em_otu_grouped <- create_grouped_matrix(
asv_matrix = em_otu,
metadata = em_metadata,
sample_col = "SampleID",
group_col= "SampleType",
group_order = "metadata"
)

em_dendrogram <- build_dendrogram(
  mat = em_otu_grouped,
  distance_method = "bray",
  cluster_method = "ward.D2"
)

em_dendrogram_plot <- plot_dendrogram(
  dend = em_dendrogram,
  metadata = em_metadata,
  label_from = "SampleType",      
  color_by = "SampleType",
  color_palette = habitat_palette,
  point_size = 2,
  orientation = "top",
  shape_by = "Habitat",
  theme_obj = theme_void() + theme(text = element_text(size = 7, color = "black"),
  legend.title = element_text(size = 7, color = "black"),)
)

em_dendrogram_plot
```

#### Combining the plots

`combine_dendrogram_alluvial()` stacks the two plots vertically and 
aligns their x-axes to the dendrogram leaf order, so the columns of 
the alluvial plot follow the same left-to-right arrangement as the 
dendrogram tips. This alignment is handled automatically via 
`leaf_order` — without it, the two plots would use independent 
orderings and the visual connection between them would be lost.


#### Fine-tuning alignment

A practical challenge when combining dendrograms with alluvial plots 
is that dendrogram tips rarely fall exactly at integer x positions — 
branches have varying widths and the outermost tips tend to drift, 
creating a misalignment between the dendrogram leaves and the alluvial 
columns beneath them. `dend_limits_left` and `dend_limits_right` 
control the x-axis limits of the dendrogram panel via 
`coord_cartesian()`, allowing precise alignment without dropping any 
data. Increasing `dend_limits_left` adds space on the left side of 
the dendrogram panel, pushing the leftmost tip further left — away 
from the first alluvial column. Increasing `dend_limits_right` reduces 
space on the right side, pushing the rightmost tip leftward — toward 
the center and away from the last alluvial column. The two parameters 
therefore behave asymmetrically: `dend_limits_left` pulls the left tip 
outward, while `dend_limits_right` pulls the right tip inward. The 
correct values depend on the number of groups and the specific 
clustering, so some manual adjustment is expected and normal. For 
vertical dendrograms, use `dend_limits_top` and `dend_limits_bottom` 
instead.

#### Legend control

The `legend` argument controls whether legends are included in the 
combined figure: `"separate"` places legends outside the plot area, 
`"omit"` removes them entirely, and `"together"` merges them into one. 
Omitting legends is useful when full manual control over placement is 
needed — for example, when using `cowplot` or `ggpubr` to arrange 
legends alongside other figure panels.

```{r alluvial-dendrogram, fig.height = 9}

p_allu4dend <- create_alluvial_plot(
  data = example_microbiome,
  tax_level = "Class",
  group_col = "SampleType",
  prepare_args = list(clean_taxonomy = TRUE),
  palette_list = c("Reds", "Purples", "BrwnYl", "Blues", "TealGrn"),
  palette_args = list(
    cmax = 65,
    luminance = c(20, 90),
    power = 1.2
  ),
  plot_args = list(
    theme_obj = theme_phylopal(),
    line_width = 0.2,
    x_axis_label = "Sample Type"
  )
) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::guides(fill = ggplot2::guide_legend(ncol = 1))


# dendrogram and alluvial with legends
dendrogram_alluplot <- combine_dendrogram_alluvial(
  alluvial_plot   = p_allu4dend +
  scale_y_continuous(expand = c(0,0), breaks = seq(0,1,0.1), limits = c(0,1))+
  ggplot2::guides(fill = guide_legend(ncol =1, title = "Class")),
  dendrogram_plot = em_dendrogram_plot +
  ggplot2::guides(color = guide_legend(ncol = 2, title = "Sample Type"), shape = guide_legend(ncol = 2)),
  dend_position   = "top",
  dend_height     = 0.15,
  strip_alluvial_x = FALSE,
  legend          = "separate",
  legend_source   = "both",       
  legend_position = "right",
  legend_rel_width = 0.75,            
  alluvial_margins    = ggplot2::margin(0, 0, 0, 0, unit = "cm"),
  dendrogram_margins    = ggplot2::margin(0, 0, 0.15, 0, unit = "cm"),
  outer_margins    = ggplot2::margin(0.2, 0.2, 0.2, 0.2, unit = "cm"),
  align = "panel",
  x_expand_zero = TRUE,
  align_x_centers = TRUE,
  leaf_order = em_dendrogram$order,
  overwrite_x_scales = TRUE,
  dend_limits_left = 0.4,  
  dend_limits_right = 0.18
) 
grid::grid.draw(dendrogram_alluplot)
```

```{r alluvial_control_legends, fig.height = 9}
dendrogram_alluplot_control <- combine_dendrogram_alluvial(
  alluvial_plot   = p_allu4dend,
  dendrogram_plot = em_dendrogram_plot,
  dend_position   = "top",
  dend_height     = 0.15,
  strip_alluvial_x = TRUE,
  legend          = "omit",
  legend_source   = "both",       
  legend_position = "right",
  legend_rel_width = 0.75,            
  alluvial_margins    = ggplot2::margin(0, 0, 0, 0, unit = "cm"),
  dendrogram_margins    = ggplot2::margin(0, 0, 0.15, 0, unit = "cm"),
  outer_margins    = ggplot2::margin(0.2, 0.2, 0.2, 0.2, unit = "cm"),
  align = "panel",
  x_expand_zero = TRUE,
  align_x_centers = TRUE,
  leaf_order = em_dendrogram$order,
  overwrite_x_scales = TRUE,
  dend_limits_left = 0.3,  
  dend_limits_right = 0.18
) 


legend_alluplot <- ggpubr::get_legend(p_allu4dend + 
                                            guides(
                                              fill = guide_legend(ncol =1, title = "Class")))


legend_dendrogram <- ggpubr::get_legend(
  em_dendrogram_plot + 
  ggplot2::guides(color = guide_legend(ncol = 1, title = "Sample Type"), shape = guide_legend(ncol = 1))+
    ggplot2::theme(legend.position = "right", 
          legend.box = "vertical",
          legend.title.position = "top",
          plot.margin = margin(0,0,0,0))
)

p_allu_full <- cowplot::plot_grid(
  cowplot::plot_grid(
    ggplotify::as.ggplot(dendrogram_alluplot_control),
    cowplot::plot_grid(
      legend_dendrogram,
      legend_alluplot,
      rel_heights = c(0.6, 1),
      rel_widths = c(1, 1),
      ncol = 1,
      align = "hv", axis = "tblr"
    ),
    rel_widths = c(1,0.6),
    ncol = 2,
    align = "hv", axis = "tblr"
  )
)

grid::grid.draw(p_allu_full)
```

#### Convenience wrapper
The whole process can be simplified using the 
`create_alluvial_dendrogram_plot()` wrapper, which runs the full 
pipeline — grouping the ASV/OTU matrix, building the dendrogram, preparing 
and classifying alluvial data, generating the palette, and combining 
the plots — in a single call. 

It takes as input the raw ASV/OTU matrix 
(`asv_matrix`, samples as columns and ASVs/OTUs as rows), a metadata data 
frame, and the long-format ASV/OTU table with pre-calculated RA 
(`alluvial_data`). 

Arguments for each internal step are passed as 
named lists: `build_dendrogram_args` and `plot_dendrogram_args` control 
the dendrogram, while `alluvial_args` accepts nested `prepare_args`, 
`classify_args`, `palette_args`, and `plot_args` forwarded to the 
respective alluvial functions. Layout parameters like 
`dend_limits_left`, `dend_limits_right`, and `legend_rel_width` are 
direct arguments rather than nested, since they are commonly adjusted. 

The function returns a named list containing all intermediate objects 
— `grouped_matrix`, `dendrogram`, `dendrogram_plot`, `alluvial`, and 
`combined_plot` — so any component can be accessed for further 
customization or export without rerunning the pipeline.

Use the wrapper for standard workflows and drop into the step-by-step approach 
when you need to modify an intermediate result.

```{r alluvial-dendrogram-wrapper, fig.height = 10}
res <- create_alluvial_dendrogram_plot(
  asv_matrix = em_otu,
  metadata = em_metadata,
  sample_col = "SampleID",
  group_col  = "SampleType",
  alluvial_data = example_microbiome,
  tax_level = "Class",
  dend_color_palette = habitat_palette,
  dend_shape_by = "Habitat",
  theme_alluvial = theme_phylopal(),
  theme_dendrogram = ggplot2::theme_void(),
  alluvial_args = list(
    return_all = TRUE,
    prepare_args = list(clean_taxonomy = TRUE),
    classify_args = list(low_abundance_threshold = 0.01),
    palette_args = list(
      palette_list = c("Reds", "Purples", "BrwnYl", "Blues", "TealGrn"),
      cmax = 65,
      luminance = c(20, 90),
      power = 1.2
    ),
    plot_args = list(
      line_width = 0.2,
      x_axis_label = "Sample Type"
    )
  ),
  post_plot_guides   = list(      # guides applied to alluvial
    fill = ggplot2::guide_legend(ncol = 1, title = "Class")
  ),    
  dend_limits_left = 0.4,  
  dend_limits_right = 0.18, 
  combine_args = list(
    legend_rel_width = 0.5,
    strip_alluvial_x = TRUE,  
    alluvial_margins = ggplot2::margin(0, 0, 0, 0, unit = "cm"),
    outer_margins    = ggplot2::margin(0.2, 0.5, 0.2, 0.2, unit = "cm") 
  )
)

p_em_alluvial_dend_wrapper <- res$combined_plot

grid::grid.draw(p_em_alluvial_dend_wrapper)
```

## Function reference

| Function | What it does |
|---|---|
| `replace_incertae_sedis_NAs()` | Clean taxonomy: normalize Incertae Sedis, propagate parent taxa |
| `process_barplot_data()` | Aggregate ASV-level RA to taxonomic level, mark low-abundance taxa |
| `prepare_alluvial_data()` | Aggregate and complete zeros for alluvial input |
| `classify_taxa_patterns()` | Classify taxa as shared/unique/mixed-abundance across groups |
| `generate_palette_hcl()` | HCL palette with optional hierarchical grouping by higher taxonomy |
| `generate_grouped_palette()` | Assign color families to groups (e.g. for facet strip colors) |
| `generate_alluvial_palette()` | Alluvial-aware palette respecting shared/unique structure |
| `add_alpha()` | Add transparency to hex colors |
| `plot_taxonomic_barplot()` | Stacked barplot with optional colored facet strips |
| `plot_alluvial()` | Alluvial/Sankey plot |
| `build_dendrogram()` | Compute Bray-Curtis dendrogram from ASV/OTU matrix |
| `plot_dendrogram()` | Plot dendrogram with metadata-colored labels |
| `combine_dendrogram_alluvial()` | Combine alluvial + dendrogram with aligned axes |
| `create_alluvial_plot()` | Full alluvial workflow wrapper |
| `create_alluvial_dendrogram_plot()` | Full alluvial + dendrogram wrapper |
| `theme_phylopal()` | Clean built-in ggplot2 theme |

---

## References
Caporaso, J.G., et al. (2011). Global patterns of 16S rRNA diversity 
at a depth of millions of sequences per sample. *PNAS*, 108, 4516--4522.

Oksanen, J., et al. (2022). vegan: Community Ecology Package. 
R package version 2.6-4. https://CRAN.R-project.org/package=vegan

Brunson, J.C. (2020). ggalluvial: Layered Grammar for Alluvial Plots.
*Journal of Open Source Software*, 5(49), 2017.

## Citation

If you use phyloPal in your research, please cite:
Slawinska MW (2025). phyloPal: Taxonomic Color Palettes and Alluvial-Dendrogram
Visualization for Microbiome Data. R package version 0.1.0.
https://github.com/mwslawinska/phyloPal

---

## License

MIT © Magdalena W. Slawinska